A New Hybrid Machine Learning Method for Stellar Parameter Inference

### 2025-03-07 paper/paper2/paper.tex
Microscopic change
Microscopic change 2
Add teff=7000K discussion

\subsection{Line Identification with \texttt{blase}}
The next step was to convert the PHOENIX subset into a physically interpretable intermediate representation: a table of spectral line properties rather than a\textbf{n} array of fluxes.
We used \texttt{blase}, which models spectral lines as Voigt profiles and tunes the profiles to mimic the original PHOENIX spectrum with back propagation.
Back propagation, put simply, is the process of moving in the autodiff gradient field to update ML model parameters (like mentioned earlier, usually against the gradient because we minimize loss functions).
Four parameters were optimized: the line center $\mu$, the log-amplitude $\ln(a)$, the Gaussian width $\sigma$, and the Lorentzian width $\gamma$.
The optimization used the Adam optimizer \textbf{\citep{adam}} with a learning rate of 0.05 over 100 epochs.
\textbf{\citealt{blase} recommends this setup as the minimum, and for a proof-of-concept implementation, we found it best to leave it as is.}
In addition, we limited two custom parameters: wing cut to 6,000 and prominence to 0.005.
Wing cut \textbf{(in pixels)} is a parameter that determines the extent of the Voigt profile to evaluate, saving computational resources by not evaluating negligible line wings.
Prominence \textbf{(in normalized flux counts)} sets a lower limit for the amplitude of detected lines, which saves resources by disregarding shallow lines, so in our case we disregard lines \textbf{with amplitude less than} 0.005.
In short, our choices for wing cut and prominence decrease the computational cost of \texttt{blase}'s cloning process at the expense of decreasing its accuracy slightly.
\texttt{blase} uses the pseudo-Voigt approximation, which saves on computational cost while remaining accurate to about 1\% \citep{pseudovoigt}.
The pseudo-Voigt approximation uses a weighted average of a Gaussian and Lorentzian as opposed to a convolution.
\texttt{blase}'s pseudo-Voigt profile implementation uses the following:
\begin{linenomath*}
\begin{subequations}
\begin{gather}
    \tilde{\mathsf{V}}_\mu(\lambda) = a\left[\eta \mathbf{L}(\lambda - \mu'; f) + (1 - \eta)\mathbf{G}(\lambda - \mu'; f)\right]\\
    \eta = \sum_{n=1}^3 \mathbf{u}_n\left(\dfrac{2\gamma}{f}\right)^n\\
    f = 32\sum_{n=0}^5 \mathbf{v}_n \left(\sqrt{2\ln(2)}\sigma\right)^{5-n}(\gamma)^n\\
    \nonumber \mathbf{u} = \mqty[1.36603 \\ -0.47719 \\ 0.11116] \quad \mathbf{v} = \mqty[1 \\ 2.69269 \\ 2.42843 \\ 4.47163 \\ 0.07842 \\ 1]
    \label{eq:equation2}
\end{gather}
\end{subequations}
\end{linenomath*}
where $\mathbf{L}$ and $\mathbf{G}$ are abbreviations for Lorentzian and Gaussian profiles, respectively.
Notice that we use $\mu'$ instead of $\mu$ in the formula.
This is because \texttt{blase} optionally allows the line center to shift slightly during optimization, and it is this shifted center which is used in computation.
The individual Voigt profiles are still indexed by $\mu$ for cross-model line identification, explained in the next section.
Once optimization was complete, the list of identified lines was saved to a `state dictionary': a common representation for pre-trained machine learning models that can be stored to disk for reuse later.
These are stored in the \texttt{.pt} file format for each of the 1,314 PHOENIX subset grid points.
The total disk space these files take up is 465 MB (382 MB when downloaded as a zip archive).
For reference, the storage space the PHOENIX subset takes up on disk is approximately 8.1 GB.
This represents a data compression factor of around 20 just by recasting the spectrum with \texttt{blase}.
\textbf{The state dictionaries are available on Zenodo at \url{https://zenodo.org/records/11246174} \citep{zenodo}.}

@@

In \autoref{fig:figure8}, we show the residuals of the inferred stellar parameters with respect to the true values.
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figure8}
    \caption{\textbf{Residual plots for the GP minimization inference algorithm, all plotted with respect to $T_\text{eff}$ (values aggregated across $\log(g)$ and [Fe/H]).
    Notice the interesting change in behavior of all residuals at around $T_\text{eff} = 7000$ K.}}
    \label{fig:figure8}
\end{figure*}
\textbf{We see that there is a distinction between the algorithm's behavior above and below $T_{\mathrm{eff}} = 7000$ K, so we define low and high temperature regimes accordingly.
It is interesting that this observed change occurs where the PHOENIX grid changes in $T_\text{eff}$ sampling from 100 K to 200 K.}

@@

\textbf{In the low-temperature regime, our median $T_\text{eff}$ residual is about 10 K, about half that of MINESweeper \citep{minesweeper} and \texttt{astroNN} \citep{leung2019}, and about ten times larger than ASPCAP \citep{ASPCAP}.
Median $\log(g)$ and [Fe/H] residuals are 0 dex, which are comparable to all three aforementioned methods, however \texttt{blase3D} is overall less consistent as the standard deviation of the residuals are about an order of magnitude larger.
At high temperatures, the median $T_\text{eff}$ residual changes drastically to about -434 K, which is significantly worse than the other methods.
$\log(g)$ becomes a few times less accurate, and both $\log(g)$ and [Fe/H] have large standard deviations, all leading to large inconsistencies and worse performance in this temperature regime.
\texttt{blase3D} at this stage does not quite reach the level of contemporary methods that can deal with more realistic data and cover much larger parameter spaces, but the fact that, in its best case, it can perform to within an order of magnitude using a drastically simpler implementation, sparks promise for future improvements.}

@@

    Users would be able to fit a greater variety of stellar spectra in many different wavelength regimes.

@@

    Future work would need to develop ways to tune these extrinsic parameters alongside stellar parameters, enabling users to optimize these frequently-observed extrinsic parameters on top of the base stellar parameters.

>>>>

\subsection{Line Identification with \texttt{blase}}
The next step was to convert the PHOENIX subset into a physically interpretable intermediate representation: a table of spectral line properties rather than a\textbf{n} array of fluxes.
We used \texttt{blase}, which models spectral lines as Voigt profiles and tunes the profiles to mimic the original PHOENIX spectrum with back propagation.
Back propagation, put simply, is the process of moving in the autodiff gradient field to update ML model parameters (like mentioned earlier, usually against the gradient because we minimize loss functions).
Four parameters were optimized: the line center $\mu$, the log-amplitude $\ln(a)$, the Gaussian width $\sigma$, and the Lorentzian width $\gamma$.
The optimization used the Adam optimizer \textbf{\citep{adam}} with a learning rate of 0.05 over 100 epochs.
\textbf{\citealt{blase} recommends this setup as the minimum, and for a proof-of-concept implementation, we found it best to leave it as is.}
In addition, we limited two custom parameters: wing cut to 6,000 and prominence to 0.005.
Wing cut \textbf{(in pixels)} is a parameter that determines the extent of the Voigt profile to evaluate, saving computational resources by not evaluating negligible line wings.
Prominence \textbf{(in normalized flux counts)} sets a lower limit for the amplitude of detected lines, which saves resources by disregarding shallow lines, so in our case we disregard lines \textbf{with amplitude less than} 0.005.
In short, our choices for wing cut and prominence decrease the computational cost of \texttt{blase}'s cloning process at the expense of decreasing its accuracy slightly.
\texttt{blase} uses the pseudo-Voigt approximation, which saves on computational cost while remaining accurate to about 1\% \citep{pseudovoigt}.
The pseudo-Voigt approximation uses a weighted average of a Gaussian and Lorentzian as opposed to a convolution.
\texttt{blase}'s pseudo-Voigt profile implementation uses the following:
\begin{linenomath*}
\begin{subequations}
\begin{gather}
    \tilde{\mathsf{V}}_\mu(\lambda) = a\left[\eta \mathbf{L}(\lambda - \mu'; f) + (1 - \eta)\mathbf{G}(\lambda - \mu'; f)\right]\\
    \eta = \sum_{n=0}^2 \mathbf{u}_{n}\left(\dfrac{2\gamma}{f}\right)^{n+1}\\
    f = 32\sum_{n=0}^5 \mathbf{v}_n \left(\sqrt{2\ln(2)}\sigma\right)^{5-n}(\gamma)^n\\
    \nonumber \mathbf{u} = \mqty[1.36603 \\ -0.47719 \\ 0.11116] \quad \mathbf{v} = \mqty[1 \\ 2.69269 \\ 2.42843 \\ 4.47163 \\ 0.07842 \\ 1]
    \label{eq:equation2}
\end{gather}
\end{subequations}
\end{linenomath*}
where $\mathbf{L}$ and $\mathbf{G}$ are abbreviations for Lorentzian and Gaussian profiles, respectively.
Notice that we use $\mu'$ instead of $\mu$ in the formula.
This is because \texttt{blase} optionally allows the line center to shift slightly during optimization, and it is this shifted center which is used in computation.
The individual Voigt profiles are still indexed by $\mu$ for cross-model line identification, explained in the next section.
Once optimization was complete, the list of identified lines was saved to a `state dictionary': a common representation for pre-trained machine learning models that can be stored to disk for reuse later.
These are stored in the \texttt{.pt} file format for each of the 1,314 PHOENIX subset grid points.
The total disk space these files take up is 465 MB (382 MB when downloaded as a zip archive).
For reference, the storage space the PHOENIX subset takes up on disk is approximately 8.1 GB.
This represents a data compression factor of around 20 just by recasting the spectrum with \texttt{blase}.
\textbf{The state dictionaries are available on Zenodo at \url{https://zenodo.org/records/11246174} \citep{zenodo}.}

@@

In \autoref{fig:figure8}, we show the residuals of the inferred stellar parameters with respect to the true values.
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figure8}
    \caption{\textbf{Residual plots for the GP minimization inference algorithm, all plotted with respect to $T_\text{eff}$ (values aggregated across $\log(g)$ and [Fe/H]).
    Notice the interesting change in behavior of all residuals at around $T_\text{eff} = 7000$ K.}}
    \label{fig:figure8}
\end{figure*}
\textbf{We see that there is a distinction between the algorithm's behavior above and below $T_{\mathrm{eff}} = 7000$ K, so we define low and high temperature regimes accordingly.
We know that as $T_\text{eff}$ increases, we detect less lines with \texttt{blase}, as shown in \autoref{fig:figure2}, and lines tend to spread out, possibly resulting in artifacts from our wing cut approximation.
Any number of physical or computational factors could explain this behavior, although the apparent sharpness of the transition leads us to postulate that it could be due to PHOENIX grid $T_\text{eff}$ sampling changing from 100 K to 200 K at this point.}

@@

\textbf{In the low-temperature regime, our median $T_\text{eff}$ residual is about 10 K, about half that of MINESweeper \citep{minesweeper} and \texttt{astroNN} \citep{leung2019}, but about ten times larger than ASPCAP \citep{ASPCAP}.
Median $\log(g)$ and [Fe/H] residuals are 0 dex, which are comparable to all three aforementioned methods, however \texttt{blase3D} is overall less consistent as the standard deviation of the residuals are about an order of magnitude larger.
At high temperatures, the median $T_\text{eff}$ residual changes drastically to about -434 K, which is significantly worse than the other methods.
$\log(g)$ becomes a few times less accurate, and both $\log(g)$ and [Fe/H] have larger standard deviations, all leading to large inconsistencies and worse performance in this temperature regime.
\texttt{blase3D} at this stage does not quite reach the level of contemporary methods that can deal with more realistic data and cover much larger parameter spaces, but the fact that, in its best case, it can perform to within an order of magnitude using a drastically simpler implementation, sparks promise for future improvements.}

@@

    \textbf{With this, u}sers would be able to fit a greater variety of stellar spectra in many different wavelength regimes.

@@

    Future work would need to develop ways to tune these extrinsic parameters alongside stellar parameters.
